Disclosure of Product System Models in Life Cycle Assessment: Achieving Transparency and Privacy

### 2018-02-10 aggregation/agg-community.tex
Revise Semantics bit to cite FAIR

To advance LCA practice, the community should embrace the notion of a software-independent PSM description and consider how to best achieve it.  The SETAC North America roadmap module for PSM description and revision \citep{Kuczenski_JLCA_2018} includes milestones toward this goal.  The disclosure framework presented here is offered as a candidate for Milestone 2.1, ``a minimal description of a PSM.''  The community should give this proposal a robust critical evaluation, and researchers should consider how to conceptualize their own models in terms of the framework.  Other milestones and cross-cutting issues are of crucial interest:
\begin{enumerate}[label=(\alph*)]
\item LCI database providers must provide their users with unambiguous, static, resolvable references to background datasets (Milestone 1.1) so that their use in a model may be documented precisely (Milestone 3.1);
\item The transformation of multi-functional processes to single-function processes should be described using the framework (Milestones 1.2 and 1.4);
\item Software makers should enable their users to automatically generate lists of foreground, background, and emission flows a model foreground (disclosure items \ref{itm:fg} through \ref{itm:em}) and to export foreground matrices (disclosure items \ref{itm:af} through \ref{itm:bf}) from a model in a concise format (Milestones 2.2 and 3.4);
\item During ISO critical review, practitioners and reviewers should strive to come to agreement on the formal structure and contents of the model being reviewed;
\item The LCA research community, including interested parties within the International Society for Industrial Ecology, the UNEP/SETAC Life Cycle Initiative, and other organizations, should collectively pursue consistent, shared definitions of quantities of measure, flows, and contexts used in LCIA.  %These definitions should be published and curated using Semantic Web techniques and practices.
LCIA method developers should immediately begin using these consensus definitions instead of their own internal lists when publishing flow characterization data, so that software providers can interpret the publications correctly.  %A characterization of a flow defined privately, into a context defined privately, cannot be interpreted meaningfully by the public.
\end{enumerate}
As these milestones are pursued and achieved, it will be possible for a standalone description of a PSM to be computed and extended by another party, regardless of the software used to construct the model originally.  This will tremendously improve the capacity for critical review and reuse of LCA models.

>>>>

To advance LCA practice, the community should embrace the notion of a software-independent PSM description and consider how to best achieve it.  The SETAC North America roadmap module for PSM description and revision \citep{Kuczenski_JLCA_2018} includes milestones toward this goal.  The disclosure framework presented here is offered as a candidate for Milestone 2.1, ``a minimal description of a PSM.''  The community should give this proposal a robust critical evaluation, and researchers should consider how to conceptualize their own models in terms of the framework.  Other milestones and cross-cutting issues are of crucial interest:
\begin{enumerate}[label=(\alph*)]
\item LCI database providers must provide their users with unambiguous, static, resolvable references to background datasets (Milestone 1.1) so that their use in a model may be documented precisely (Milestone 3.1);
\item The transformation of multi-functional processes to single-function processes should be described using the framework (Milestones 1.2 and 1.4);
\item Software makers should enable their users to automatically generate lists of foreground, background, and emission flows a model foreground (disclosure items \ref{itm:fg} through \ref{itm:em}) and to export foreground matrices (disclosure items \ref{itm:af} through \ref{itm:bf}) from a model in a concise format (Milestones 2.2 and 3.4);
\item During ISO critical review, practitioners and reviewers should strive to come to agreement on the formal structure and contents of the model being reviewed;
\item The LCA research community, including interested parties within the International Society for Industrial Ecology, the UNEP/SETAC Life Cycle Initiative, and other organizations, should collectively pursue consistent, shared definitions of quantities of measure, flows, and contexts used in LCIA.  %These definitions should be published and curated using Semantic Web techniques and practices.
LCIA method developers should immediately begin using these consensus definitions instead of their own internal lists when publishing flow characterization data, so that software makers can interpret the publications correctly.  %A characterization of a flow defined privately, into a context defined privately, cannot be interpreted meaningfully by the public.
\end{enumerate}
As these milestones are pursued and achieved, it will be possible for a standalone description of a PSM to be computed and extended by another party, regardless of the software used to construct the model originally.  This will tremendously improve the capacity for critical review and reuse of LCA models.

### 2018-02-10 aggregation/agg-semantic.tex
Revise Semantics bit to cite FAIR

In practice, while it is easy to reproduce a set of sparse matrices, there is considerable potential for ambiguity
in identifying the background datasets and emissions to which those matrix entries correspond.
In order to achieve the transparency and reproducibility promised by this disclosure framework, it must be possible for study authors and readers to agree on the meaning of those references.  In particular, the database configuration, dataset version, reference flow and dimension (quantitative unit of measure) must be unambiguous and easy to interpret.

Agreement about the meaning of concepts is often accomplished using linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.  The resource at the end of the link functions as a point of agreement: both study author and data user can follow it from anywhere on the Internet to obtain the same information.  The content pointed to by the hyperlink, along with context provided by \emph{other} references to the same content, gives meaning to the information and allows it to be curated by the study authors and other members of the community \citep{Khan_2011}.   %Semantic data are structured by reference to knowledge models called ontologies, which describe the relationships among different types of entities \citep{Madin2008}.  Some preliminary descriptions of the entity types involved in LCA computation, and their relationships to one another, have recently been developed \citep{Ciroth_Srocka_2014, Janowicz_WOP_2015, Kuczenski_JCP_2016}.

Before the goal of easily reproducible foreground models can be realized, this linked data foundation must be laid.  Background database providers should ensure that their users have access to \emph{stable semantic references} for publicly available datasets that denote a particular activity in a particular database version and configuration.  These references should take the form of URIs that can be accessed using any Internet browser and can provide both author and reader with documentary information describing the referred dataset, the available reference flows and their dimensions, and a mechanism for the reader to obtain access to the exchange data or LCI/LCIA computations for the purposes of model validation.  A dependency reference in disclosure item~\ref{itm:bg} should include both an activity and a reference flow.

>>>>

In practice, while it is easy to reproduce a set of sparse matrices, there is considerable potential for ambiguity in identifying the background datasets and emissions to which those matrix entries correspond.
In order to achieve the transparency and reproducibility promised by this disclosure framework, it must be possible for study authors and readers to agree on the meaning of the references contained in items~\ref{itm:bg} and \ref{itm:em}.
The FAIR guidelines, which were designed to help scientists organize data for Findability, Accessibility, Interoperability, and Reusability \citep{Wilkinson_2016}, provide valuable perspective on this goal.

Central to many of the FAIR guiding principles is the use of linked semantic data \citep{Bizer_2009}, in which an object to be interpreted is signified by a link to a resource on the World Wide Web, typically referred to as a Uniform Resource Identifier (URI) or a hyperlink.  This link serves as a unique identifier and a resolvable reference to metadata, and the linked content can be  curated by the study authors and other members of the community \citep{Khan_2011}.

%Semantic data are structured by reference to knowledge models called ontologies, which describe the relationships among different types of entities \citep{Madin2008}.  Some preliminary descriptions of the entity types involved in LCA computation, and their relationships to one another, have recently been developed \citep{Ciroth_Srocka_2014, Janowicz_WOP_2015, Kuczenski_JCP_2016}.

Before the goal of easily reproducible foreground models can be realized, this linked data foundation must be laid.  Background database providers should ensure that their users have access to \emph{stable semantic references} for publicly available datasets that denote a particular activity in a particular database version and configuration.  These references should take the form of URIs that can be accessed using any Internet browser and can provide both author and reader with documentary information describing the referred dataset, including version information; database configuration; the available reference flows and their dimensions (quantitative units of measure); and a mechanism for the reader to obtain access to the exchange data or LCI/LCIA computations for the purposes of model validation.  A dependency reference in disclosure item~\ref{itm:bg} should include both an activity and a reference flow.
